# RStudio Exercise 3 - Analysis


Read the learning2014 data from my local folder into RStudio

```{r}
alc <- read.table("H:/Jatko-opinnot/Open data science/IODS-project/data/alc", sep=",", header=TRUE)
```


Explore the column names and dimensions of the data

```{r}
colnames(alc)
dim(alc)
```

Data has information about student alcohol consumption. The original data set can be found here:  https://archive.ics.uci.edu/ml/datasets/STUDENT+ALCOHOL+CONSUMPTION

There are 382 observations (rows) and 36 variables (columns) in the data. The variables include information about alcohol consumption and also social, gender and study time attributes.


I chose the following 4 variables to study the relationship between them and alcohol consumption level:

14 studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours) 
19 activities - extra-curricular activities (binary: yes or no) 
29 health - current health status (numeric: from 1 - very bad to 5 - very good)
30 absences - number of school absences (numeric: from 0 to 93) 


My hypothesis:
High weekly study time is associated with low alcohol consumption
Having extra-curricular activities is associated with low alcohol consumption
Better health is associated with low alcohol consumption
The higher the number of absences the larger proportion of students have high alcohol consumption

Tabulate the target variable versus the alcohol consumption
```{r}
table(high_use = alc$high_use, studytime = alc$studytime)
table(high_use = alc$high_use, activities = alc$activities)
table(high_use = alc$high_use, health = alc$health)
table(high_use = alc$high_use, absences = alc$absences)
```

Create bar plots for each variable 

First access the necesary libraries
```{r}
library(tidyr); library(dplyr); library(ggplot2)

```
 Then initialize a plot of 'studytime'
```{r}
g1 <- ggplot(alc, aes(studytime))
```

And draw a bar plot of high_use by studytime
```{r}
g1 + facet_wrap("high_use") + geom_bar()
```

Do the same for activities, health and absences
```{r}
g2 <- ggplot(alc, aes(activities))
g2 + facet_wrap("high_use") + geom_bar()

g3 <- ggplot(alc, aes(health))
g3 + facet_wrap("high_use") + geom_bar()

g4 <- ggplot(alc, aes(absences))
g4 + facet_wrap("high_use") + geom_bar()
```


Create box plots

```{r}
g5 <- ggplot(alc, aes(x = high_use, y = studytime))
g5 + geom_boxplot() + ylab("studytime")

g6 <- ggplot(alc, aes(x = high_use, y = activities))
g6 + geom_boxplot() + ylab("activities")

g7 <- ggplot(alc, aes(x = high_use, y = health))
g7 + geom_boxplot() + ylab("health")

g8 <- ggplot(alc, aes(x = high_use, y = absences))
g8 + geom_boxplot() + ylab("absences")
```

My hypothesis:
High weekly study time is associated with low alcohol consumption
Result: 
True: those who had low alcohol consumption studied more.

My hypothesis:
Having extra-curricular activities is associated with low alcohol consumption
Result:
The activities doesn't seem to affect the level of alcohol consumption

My hypothesis:
Better health is associated with low alcohol consumption
Result:
Health doesn't seem to affect the level of alcohol consumption 

My hypothesis:
The higher the number of absences the larger proportion of students have high alcohol consumption
Result:
True: Students with low alcohol consumption has fewer absences

Logistic regression to statistically explore the relationship between alcohol consumption level and the target variables

Find the model with glm()
```{r}

m <- glm(high_use ~ studytime + activities + health +absences, data = alc, family = "binomial")
```

Print out a summary of the model
```{r}
summary(m)
```
The p-values tell that only studytime and absences have effect on alcohol consumption level. This differs from my hypothesis where I thought that also activities and health would have effect on alcohol consumption level. 


Compute the coefficients as odds ratios (OR)
```{r}
OR <- coef(m) %>% exp
```

Compute confidence intervals (CI)
```{r}
CI <- confint(m) %>% exp
```

Present the odds ratios with their confidence intervals
```{r}
cbind(OR, CI)
```


