# RStudio Exercise 2 - Analysis


Read the learning2014 data from my local folder into RStudio

```{r}
learning2014 <- read.table("H:/Jatko-opinnot/Open data science/IODS-project/data/learning2014", sep=",", header=TRUE)
```


Explore the dimensions and structure of the data

```{r}
dim(learning2014)
str(learning2014)
```

Data is from international survey of Approaches to Learning and data is collected 3.12.2014 - 10.1.2015

There are 166 observations (rows) and 8 variables (columns). 
First column has row numbers as integer. Columns "age" and "points" are also as integers and columns "attitude", "deep", "stra" and "surf" as numerical. "Gender" is a factor with 2 levels.

Variables:
gender: gender of respondent, M = male, F = female
age: age of the respondent
attitude: global attitude towards statistics
deep: average of answers to questions related to deep learning
stra: average of answers to questions related to strategic learning
surf: average of answers to questions related to surface learning
points: exam points


Remove the first column from the data because it is not needed. 
First choose the columns to keep

```{r}
library(dplyr)
keep_columns <- c("gender","age","attitude", "deep", "stra", "surf", "points")
```

And then select the "keep_columns" to a new dataset

```{r}
learning2014 <- select(learning2014, one_of(keep_columns))
```


Graphical overview of the data
[-1] excludes the first column with row numbers. col=gender sets colors by gender

```{r}
library(GGally)
library(ggplot2)

ggpairs(learning2014, mapping = aes(col=gender, alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))
```


Summaries of the variables
```{r}
summary(learning2014$gender)
summary(learning2014$age)
summary(learning2014$attitude)
summary(learning2014$deep)
summary(learning2014$stra)
summary(learning2014$surf)
summary(learning2014$points)

```

There are twice as many women than man in the data. Youngest student in 17 and oldets 55 mean age being 25.5. Men are in average older than women. Mens attitude is on higher level than womens. On deep questions men have given sligthly bigger values than women. On strategic and surface questions women have given bigger values than men. There are hardly any difference in points between men and women. 

Highest positive correlation can be found between points and attitude (0.437). This means that attitude explains the points the student gets by 44%. Highest negative correalation is between surface and deep question. 

All variables are quite normally distributed except age where most students are quite young (3/4 under 27 years old). 


I choose variables "attitude", "strategic questions (stra)" and "surface questions (surf)" as explanatory variables to explain the exam points and fit a regression model

```{r}
mymodel <- lm (points ~ attitude + stra + surf, data = learning2014)
summary(mymodel)
```

Residuals seem quite evenly distributed with median close to zero and first and third quantile almost evenly far from the median. Minimun value is almost twice as far from median than the maximum value but we can still say that residuals seem ok.

The estimate for "attitude" tells that when attitude increses 1 unit the points increase by 3.3952 units. p-value is very small, so we can say with high confidense that attitude effects the points. For strategic questions the estimate is 0.8531 and for surface questions -0.5861 but for these varaibles the p-values are quite large (0.11716 and 0.46563) so these varaibles don't have an effect on points. 

Multiple R-squared tells that 20.74% of variability in points is explained by these three variables. 


Because variables stra and surf do not have a statistically significant relationship with the points variable these are removed from the model

```{r}
mymodel2 <- lm (points ~ attitude, data = learning2014)
summary(mymodel2)
```

Again, residuals seen to be ok.

Estimate for attitude is now a bit higher (3.5255) than in the previous model and p-value is even smaller. 

Multiple R-squared is almost as hig as before (19.06). 

Excluding the stra and surf variables from the model doesn't really change the goodnes of the model so the variables can be left out.

Diagnostic plots: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage

```{r}
par(mfrow = c(2,2)) 
plot(mymodel2, which = c(1, 2, 5))
```

Assumptions of linear model:
.The errors are normally distributed
.The errors are not correlated
.The errors have constant variance, ??2
.The size of a given error does not depend on the explanatory variables


Residual vs Fitted values: Size of an error is not dependent on fitted values --> model is ok

Normal QQ-plot: Errors of the model are nicely normally distributed --> model is ok

Residuals vs Leverage: None observation has an unusually high impact --> model is ok